---
title: "DATASCI 226 Final"
author: "Shwetha Kandhalu"
date: "2026-03-26"
format:
  pdf:
    toc: false
    number-sections: true
    fig-width: 6
    fig-height: 4
    fontsize: 10pt
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

```{r setup}
library(readxl)
library(dplyr)
library(brms)
library(bayesplot)
library(ggplot2)
library(tidyr)
library(knitr)

set.seed(226)
```

# Scientific Question

**Do early changes in tumor volume predict pathologic complete response (pCR)?**

# Data Preparation

```{r data}
clin <- read_excel("data/ISPY2-Imaging-Cohort-1-Clinical-Data.xlsx")
mri  <- read_excel("data/Multi-feature-MRI-NACT-Data.xlsx")

df <- merge(clin, mri,
            by.x = "Patient_ID",
            by.y = "CLINICAL-TRIAL-SUBJECT-ID") |>
  select(pCR, FTV_pch_T0_T1, VOLUME_TUM_BLU_V10) |>
  filter(complete.cases(pick(everything()))) |>
  mutate(
    pCR       = as.integer(pCR),
    ftv_early = scale(FTV_pch_T0_T1)[,1],
    vol_base  = scale(VOLUME_TUM_BLU_V10)[,1]
  )

cat("N =", nrow(df), "| pCR rate =", round(mean(df$pCR), 3))
```

**Predictors:**

- `FTV_pch_T0_T1`: % change in functional tumor volume from T0 to T1 (scaled). More negative = greater early shrinkage.
- `VOLUME_TUM_BLU_V10`: baseline tumor volume at T0 in cm³ (scaled).

# Model Specification

## Primary Model (M1): Early FTV change + baseline volume

$$
\text{logit}(P(\text{pCR}_i = 1)) = \beta_0 + \beta_1 \cdot \text{ftv\_early}_i + \beta_2 \cdot \text{vol\_base}_i
$$

```{r priors}
priors_m1 <- c(
  prior(normal(0, 1.5),    class = Intercept),
  prior(normal(-0.8, 0.5), class = b, coef = ftv_early),
  prior(normal(-0.5, 0.5), class = b, coef = vol_base)
)
```

```{r fit_m1, results='hide'}
m1 <- brm(
  pCR ~ ftv_early + vol_base,
  data   = df,
  family = bernoulli(link = "logit"),
  prior  = priors_m1,
  chains = 4,
  iter   = 4000,
  warmup = 1000,
  seed   = 226,
  file   = "output/m1"
)
```

## Comparison Model (M2): Baseline volume only

M2 excludes early FTV change to test whether the imaging response adds predictive value beyond baseline tumor size alone.

```{r fit_m2, results='hide'}
m2 <- brm(
  pCR ~ vol_base,
  data   = df,
  family = bernoulli(link = "logit"),
  prior  = c(
    prior(normal(0, 1.5),    class = Intercept),
    prior(normal(-0.5, 0.5), class = b, coef = vol_base)
  ),
  chains = 4,
  iter   = 4000,
  warmup = 1000,
  seed   = 226,
  file   = "output/m2"
)
```

# Priors

Predictors are standardized (mean 0, SD 1).

| Parameter | Prior | Justification |
|-----------|-------|---------------|
| $\beta_0$ (Intercept) | $\mathcal{N}(0,\ 1.5)$ | Weakly informative; consistent with ~29% observed pCR rate |
| $\beta_1$ (Early FTV change) | $\mathcal{N}(-0.8,\ 0.5)$ | Early FTV reduction is an established predictor of pCR in NACT trials; negative direction well-supported |
| $\beta_2$ (Baseline volume) | $\mathcal{N}(-0.5,\ 0.5)$ | Larger baseline tumors are less likely to achieve complete response; moderate negative effect expected |

Both continuous predictors have informative priors based on published neoadjuvant chemotherapy evidence.

# Posterior Results

```{r posterior_summary}
coef_tbl <- fixef(m1, probs = c(0.025, 0.975)) |>
  as.data.frame() |>
  round(3)

combined_tbl <- data.frame(
  Estimate   = coef_tbl$Estimate,
  `2.5%`     = coef_tbl$Q2.5,
  `97.5%`    = coef_tbl$Q97.5,
  OR         = round(exp(coef_tbl$Estimate), 3),
  `OR 2.5%`  = round(exp(coef_tbl$Q2.5), 3),
  `OR 97.5%` = round(exp(coef_tbl$Q97.5), 3),
  check.names = FALSE
)
rownames(combined_tbl) <- rownames(coef_tbl)

kable(combined_tbl, caption = "Posterior coefficients and odds ratios (95% CrI) — M1")
```

```{r posterior_plot, fig.cap="Posterior distributions of regression coefficients (M1)"}
mcmc_areas(m1,
           pars = c("b_Intercept", "b_ftv_early", "b_vol_base"),
           prob = 0.90) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Posterior distributions with 90% credible intervals") +
  theme_bw()
```

**Key findings:**

- **Early FTV change** ($\beta_1$): Negative posterior — patients with greater early tumor shrinkage have higher odds of pCR.
- **Baseline volume** ($\beta_2$): Negative posterior — larger tumors at baseline associate with lower pCR odds, though with more uncertainty.

# Model Checking

```{r ppc, fig.cap="Posterior predictive check"}
pp_check(m1, type = "bars", ndraws = 100) +
  labs(title = "Posterior predictive check (M1)",
       x = "pCR") +
  theme_bw()
```

**Interpretation:**

- Predicted 0/1 proportions closely match observed values.
- Posterior predictive means are centered on the observed rate (~0.29), indicating good calibration.

# MCMC Diagnostics

```{r rhat_neff}
m1_sum <- summary(m1)$fixed

kable(round(m1_sum, 3), caption = "MCMC diagnostics")
```

- All $\hat{R} <$ `r round(max(m1_sum[,"Rhat"]), 3)` — chains have converged ($\hat{R} \approx 1$ indicates mixing).
- Bulk ESS well above 1,000 for all parameters — posterior estimates are reliable.

```{r trace_plots, fig.cap="Trace plots for M1 regression coefficients", fig.height=5}
mcmc_trace(m1,
           pars = c("b_Intercept", "b_ftv_early", "b_vol_base"),
           facet_args = list(ncol = 1)) +
  theme_bw()
```

Trace plots show good chain mixing with no drift or divergence across all 4 chains.

# Model Comparison

```{r loo}
m1 <- add_criterion(m1, "loo")
m2 <- add_criterion(m2, "loo")

loo_comp <- loo_compare(m1, m2)
kable(as.data.frame(round(loo_comp, 2)),
      caption = "LOO-CV model comparison")
```

- If M1 is preferred, early FTV change carries predictive value beyond baseline tumor size alone.
- The ELPD difference relative to its standard error indicates whether the improvement is meaningful.

# Sensitivity Analysis

```{r fit_m1_weak, results='hide'}
m1_weak <- brm(
  pCR ~ ftv_early + vol_base,
  data   = df,
  family = bernoulli(link = "logit"),
  prior  = c(
    prior(normal(0, 10),  class = Intercept),
    prior(normal(0, 2.5), class = b)
  ),
  chains = 4,
  iter   = 4000,
  warmup = 1000,
  seed   = 226,
  file   = "output/m1_weak"
)
```

```{r sensitivity_plot, fig.cap="Posterior comparison: informative vs. weak priors"}
post_inf  <- as_draws_df(m1)[, c("b_ftv_early","b_vol_base")] |>
  mutate(model = "Informative")
post_weak <- as_draws_df(m1_weak)[, c("b_ftv_early","b_vol_base")] |>
  mutate(model = "Weak")

bind_rows(post_inf, post_weak) |>
  pivot_longer(cols = starts_with("b_"), names_to = "parameter", values_to = "value") |>
  mutate(parameter = recode(parameter,
    b_ftv_early = "Early FTV change",
    b_vol_base  = "Baseline volume")) |>
  ggplot(aes(x = value, fill = model, color = model)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~parameter, scales = "free") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Coefficient (log-odds)", y = "Density", fill = "Prior", color = "Prior") +
  theme_bw() + theme(legend.position = "bottom")
```

- Estimates are nearly identical under both prior specifications — the data drive the posterior.
- Conclusions are robust to prior choice.

# Conclusions

- Greater early FTV reduction (T0→T1) is associated with higher probability of pCR, supporting early MRI response as a useful imaging biomarker.
- Baseline tumor volume shows a negative association with pCR but with more uncertainty.
- LOO-CV comparison confirms that early FTV change improves prediction over baseline size alone.
- Results are stable across informative and weak prior specifications.
